---
title: "Class 12: IV II and RDD II"
author: "Andrew Heiss"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```

```{r packages-data, include=FALSE}
library(tidyverse)
library(kableExtra)
library(scales)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-3

# In-person<br>session 12

.class-info[

**April 6, 2023**

.light[PMAP 8521: Program evaluation<br>
Andrew Young School of Policy Studies
]

]

---

name: outline
class: title title-inv-8

# Plan for today

--

.box-1.medium[Exam 2]

--

.box-3.medium[Synthetic data fun times!]

---

layout: false
name: faqs
class: center middle section-title section-title-1 animated fadeIn

# Exam 2

---

layout: false
name: synthetic-data
class: center middle section-title section-title-3 animated fadeIn

# Synthetic data<br>fun times!

---

class: title title-3

# Basic process

--

.box-inv-3.medium[1: Draw a DAG]

--

.box-inv-3.medium[2: Create standalone exogenous columns]

--

.box-inv-3.medium[3: Connect endogenous columns]

--

.box-inv-3.medium[4: Polish columns]

--

.box-3.large.sp-before-half[Iterate. Iterate so so much.]
