---
title: "Theories of change and logic models"
date: "2023-01-26"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sHjKJRcD-ebQUbcaFhc0cQH"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "dwsCSqczNng", "Introduction", "1", TRUE,
  "XkHUQJp5fh0", "Reproducibility", "reproducibility", FALSE,
  "Uos8RumBrzE", "Program theories", "program-theories", FALSE,
  "8CJ7M2dTmF4", "Logic models & results chains", "logic-models", FALSE
)
```

## Readings

- {{< fa book >}} Chapter 2 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapter 2 in *Evaluation: A Systematic Approach* [@RossiLipseyHenry:2019]. This is available on iCollege.
- {{< fa book >}} Chapter 3 in *Evaluation: A Systematic Approach* [@RossiLipseyHenry:2019]. This is available on iCollege.


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/03-slides")
slide_tabs(slide_details, "/slides/03-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Session 3 FAQ slides**](/slides/03-class.html) ([PDF](/slides/03-class.pdf))
- [Errors vs. warnings vs. messages](https://moderndive.com/1-getting-started.html#messages) (i.e. what to do when R shows you red text)
- R Markdown examples:
    - [Quarto!](https://quarto.org/)
    - Example R Markdown file used as a code-through or step-by-step teaching document:
        - Lots of [blog posts here](https://www.andrewheiss.com/blog/)
        - Julia Silge, ["Modeling human/computer interactions on Star Trek from #TidyTuesday with workflowsets"](https://juliasilge.com/blog/star-trek/)
        - Bob Rudis, ["Some Covid Donuts To End The Week"](https://rud.is/b/2021/08/13/some-covid-donuts-to-end-the-week/)
        - Holger K. von Jouanne-Diedrich, ["The “Youth Bulge” of Afghanistan: The Hidden Force behind Political Instability"](https://blog.ephorie.de/the-youth-bulge-of-afghanistan-the-hidden-force-behind-political-instability)
    - Example R Markdown file used as a publicly-consumable document: 
        - Click on the "Manuscript" menu item [at this site](https://stats.andrewheiss.com/cautioning-canary/)
        - See [the Rmd file here](https://github.com/andrewheiss/cautioning-canary/blob/master/manuscript/manuscript.Rmd)

Hands-on R materials:

- [{{< fa brands r-project >}} Introduction to the tidyverse Posit.cloud project](https://posit.cloud/spaces/323551/content/5254772)
- [{{< fa file-zipper >}} Introduction to the tidyverse project `.zip` file](/projects/01-class.zip)
- [{{< fa brands r-project >}} Regression playground Posit.cloud project](https://posit.cloud/spaces/323551/content/5299495)
- [{{< fa file-zipper >}} Regression playground project `.zip` file](/projects/regression-example.zip)

- [Lab slides 1: Markdown and universal writing](/slides/01-class_01_markdown-writing.html) ([PDF](/slides/01-class_01_markdown-writing.pdf))
- [Lab slides 2: Getting started with R and RStudio](/slides/01-class_02-getting-started.html) ([PDF](/slides/01-class_02-getting-started.pdf))
- [Lab slides 3: Data basics](/slides/01-class_03_data-basics.html) ([PDF](/slides/01-class_03_data-basics.pdf))
- [Lab slides 4: Visualize data with ggplot2](/slides/01-class_04_visualize-data.html) ([PDF](/slides/01-class_04_visualize-data.pdf))
- [Lab slides 5: Transform data with dplyr](/slides/01-class_05_transform-data.html) ([PDF](/slides/01-class_05_transform-data.pdf))


## Bayesian statistics resources

In class I briefly mentioned the difference between frequentist and Bayesian statistics. You can see a bunch of [additional resources and examples of these two approaches to statistics here](/resource/bayes.qmd). [This huge blog post](https://www.andrewheiss.com/blog/2021/12/01/multilevel-models-panel-data-guide/) also shows how to do multilevel models with Bayesian models.


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sHjKJRcD-ebQUbcaFhc0cQH

https://youtu.be/dwsCSqczNng
PMAP 8521 • (3) Theories of change: (0) Introduction

Introduction to the class session on theories of change

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/


https://youtu.be/XkHUQJp5fh0
PMAP 8521 • (3) Theories of change: (1) Reproducibility

Discussion about the importance of reproducible data analysis practices

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/

  
https://youtu.be/Uos8RumBrzE
PMAP 8521 • (3) Theories of change: (2) Program theories

Discussion of program components, program impact theories, and the importance of basing programs on sound theories

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/


https://youtu.be/8CJ7M2dTmF4
PMAP 8521 • (3) Theories of change: (3) Logic models & results chains

Discussion of results chains, logic models, and the difference between logic models and impact theories 

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/03-content/
```
